---
title: "Green Turtle Raw Data Summary"
author: "Josh Cullen"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width=7,
  fig.retina=2,
  fig.align = "center"
  )
```

# Background

This is a brief document summarizing the green turtle (*Chelonia mydas*) satellite telemetry data that I've received from each of my collaborators (Mariana Fuentes, Meg Lamont, Chris Marshall, Camila Domit, and Chris Sasso). The main emphasis is to provide a general perspective of where these relocations are found in space by region (**Gulf of Mexico**, **Brazil**, **Qatar**) as well as by age class (**Juvenile**, **Adult**). Please let me know if anything stands out as incorrect (namely the age classification for some of these different regions).


# Summary

```{r load}

library(tidyverse)
library(lubridate)
library(sf)
library(rnaturalearth)
library(bayesmove)
library(wesanderson)


### Load data and check column names ###

dat<- list.files(path = "../Raw_data/",
             pattern = "(Cm)",
             full.names = T) %>%
  map(~read.csv(.))

cols<- c('Ptt', 'Date', 'Longitude', 'Latitude', 'Type', 'Quality', 'Error.radius',
         'Error.Semi.major.axis', 'Error.Semi.minor.axis', 'Error.Ellipse.orientation')

# Add 'Type' column to Meg Lamont's and Chris Marshall's data (elements 4 and 6)
dat[c(4,6)]<- dat[c(4,6)] %>%
  map(., ~mutate(., Type = "Argos"))


# Convert all Ptt cols to character and bind all list elements together
dat2<- dat %>%
  map(., ~mutate(., across(Ptt, as.character))) %>%
  map_df(., dplyr::select, all_of(cols))  #select only pertinent columns


# Load in metadata for tags
turt.meta<- read.csv("../Raw_data/Turtle tag metadata.csv")


# Join metadata w/ turtle obs
dat3<- left_join(turt.meta, dat2, by = 'Ptt') %>%
  relocate(Ptt, .before = Date)
```


```{r clean_and_prep}

dat3$Date<- as_datetime(dat3$Date)
dat3$Quality<- ifelse(dat3$Type == "FastGPS", NA, dat3$Quality)
```


```{r table}

# Look at no. of obs by region and age class (before filtering by behavioral state)
dat3 %>%
  group_by(Region, Age) %>%
  tally() %>% 
  knitr::kable()
```



```{r load_spatial}

### Create summary maps per region and age class ###

states<- ne_states(country = 'United States of America', returnclass = 'sf') %>%
  filter(!name %in% c('Alaska', 'Hawaii'))

north.am<- ne_countries(scale = 10, continent = "North America", returnclass = 'sf')
south.am<- ne_countries(scale = 10, continent = "South America", returnclass = 'sf')
asia<- ne_countries(scale = 10, continent = "Asia", returnclass = 'sf')
world<- ne_countries(scale = 110, returnclass = 'sf')

pal<- wes_palette('Zissou1', n = 5)[c(1,5)]

```



# Gulf of Mexico
```{r GoM_map}

ggplot() +
  geom_sf(data = north.am) +
  geom_sf(data = states) +
  geom_point(data = dat3 %>%
               filter(Region == 'GoM'), aes(Longitude, Latitude, color = Age),
             alpha = 0.3) +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 18),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_sf(xlim = c(-97, -76), ylim = c(18, 31))
```


# Brazil
```{r brazil_map}

ggplot() +
  geom_sf(data = south.am %>%
            filter(name == 'Brazil')) +
  geom_point(data = dat3 %>%
               filter(Region == 'Brazil'), aes(Longitude, Latitude, color = Age),
             alpha = 0.1) +
  annotate(geom = "text", label = "italic(Brasil)", x = -45, y = -10, size = 8,
           parse = TRUE) +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 18),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_sf(xlim = c(-50, -30), ylim = c(-30, 0))
```


# Qatar
```{r qatar_map}

ggplot() +
  geom_sf(data = asia %>%
            filter(name %in% c('Qatar','Saudi Arabia','United Arab Emirates','Bahrain',
                               'Iran'))) +
  geom_point(data = dat3 %>%
               filter(Region == 'Qatar'), aes(Longitude, Latitude, color = Age),
             alpha = 0.3) +
  annotate(geom = "text", label = "italic(Qatar)", x = 51.15, y = 25, size = 6,
           parse = TRUE) +
  annotate(geom = "text", label = "italic(KSA)", x = 50.5, y = 24.25, size = 6,
           parse = TRUE) +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 18),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_sf(xlim = c(50, 54), ylim = c(24, 27))
```


# Global view

```{r global_map}

# Global View
dat3.sf<- dat3 %>% 
  st_as_sf(., coords = c('Longitude','Latitude'), crs = 4326)

ggplot() +
  geom_sf(data = world) +
  geom_sf(data = dat3.sf, aes(color = Age), alpha = 0.3) +
  scale_color_manual(values = pal) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 18),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_sf(crs = st_crs("+proj=moll"))
```



Ultimately, I'll be using a state-space model to provide better estimates of the locations (for Argos-based positions) and then estimate behavioral states to only retain the resident/foraging/ARS state for further analysis within the species distribution models. I will be building the model using the Gulf of Mexico (GoM) data and will then validate the model using the datasets from Brazil and Qatar.
